Aggressive Scheduling for Numerical Programs

### 2012-05-02 report/draft.tex
A bunch of fixes.

\begin{abstract}
   Certain classes of algorithms are still hard to optimize by modern compilers
   due to the differences between computer architectures.
   Numerical algorithms, for example, perform better when taking advantage of features
   such as software pipelining and loop unrolling. Expert programmers are usually able to
   get the last drop of performance from the hardware, however, compilers cannot perform the same
   feat for loss of generality. The SPIRAL system bridges this gap by generating machine specific code
   from high level algorithmic descriptions. We used the SPIRAL system to implement software
   pipelining for numerical algorithms with XXX (talk about scheduling). Our results show that
   by unrolling and applying software pipelining at the C code level, we can
   improve performance by up to 30\% over unrolling alone.
\end{abstract}

@@

In this project we explore the problem of scheduling code at the high-level and generate high-performance code that is comparable
to an expert. XXX

@@

\begin{description}
   \item[Software Pipelining] We developed a transform that takes a numerical algorithm and adds software pipelining support
   using three stages. We are able to use this optimization along with other optimizations such as loop unrolling and loop tilling.
   Moreover, we were able to apply a technique that disallows C compilers to perform their own scheduling to the code. At the same
   time, we still use all the other optimizations provided by the C compiler. 
   \item[Evaluation] We evaluated our programs using several tunable parameters for software pipelining,
   loop unrolling and loop tilling. The results XXX
\end{description}

@@

Software pipelining is a coding technique that optimizes loops by overlapping instructions
from different loop iterations in order to exploit instruction level parallelism.
Due to this overlap we say that software pipelining is a type of out-of-order execution.
While software pipelining is somewhat similar to hardware pipelining, the order of the executions
is not determined by the processor but by the compiler or the programmer.

@@

Most modern compilers do not preserve instruction ordering imposed at the
input language level, and instead rely on general purpose methods such
as \emph{list scheduling} to order instructions. In the common case, this frees
the developer from the task of determining an efficient schedule, but in our
case this feature serves as a roadblock. While the GCC~\cite{gccwebsite} provides flags for
disabling instruction scheduling it comes at the cost of overall
performance because we only plan to schedule loop code. Moreover, our plan
involves using the Intel Compiler~\cite{iccwebsite}, thus we need a general solution for
scheduling at the C code level.

@@

Through the ability to apply software pipelining on any loop in SPIRAL and a
mechanism to enforce scheduling we can tailor a schedule of any
operation written in ICode to a given system. Currently, our optimizations for
tailoring software pipelining include strip-mining, reordering and unrolling the
outer loops before we perform the actual scheduling. These few optimization
provide a great degree of freedom with respect to tuning an implementation to
a system.

However, our current system is not without limitations. Because we are
attempting to schedule at a high level, our ICode implementation must map to the
underlying Instruction Set Architecture(ISA). Furthermore, our method
restricts the quality of optimizations that the compiler can provide, thus the
ICode must be aggressively optimized. Fortunately, SPIRAL was designed with
the necessary facilities in place for this level of optimization.

@@

The goal of our experiments is to demonstrate the effectiveness of software
pipelining at the C code level for automatically generated numerical
operations. Instruction scheduling is one of the few optimizations that expert
has over SPIRAL. Thus, showing that software pipelining in addition to loop
unrolling outperforms unrolling alone will help in bridging the gap between
experts and APGS like SPIRAL.

%% NOTE: need to explain this WAY better in the future.
For our experiments we benchmarked multiple automatically generated variants
of Matrix-Vector(GEMV) on several systems with different
microarchitectures. The variants included a baseline in which no loops were
unrolled, two unrolled variants where the inner-most loop was fully unrolled
and the other dimension was unrolled by either 1 or 4. Additionally, there
were two variants that were software pipelined versions of the unrolled
code. The two different unrollings resulted in completely different schedules
in the software pipelined code. It is worth noting that we specifically chose
our input sizes to fit entirely in the L1 cache, so that performance would
mostly be impacted by the number of instruction stalls. Lastly, we were using
4-way 32bit floating point vector instructions.

%% NOTE: in future, need actual parameters on these systems.
The systems we tested on were selected for their varied scheduling
facilities with respect to instruction window size where applicable. We chose three
out-of-order architectures (Core i7 960, Core2 X9650, Xeon E5320) and a single
in-order system (Intel Atom).

@@

As anticipated, the use of software pipelining provided a substantial
performance gain over unrolling alone, and on the more modern system (Core i7,
Fig.~\ref{fig:gemv_corei7}) yielded an impressive 1.8x speedup over unrolling
alone. The other two out-of-order systems (Fig.~\ref{fig:gemv_core2} and
Fig.~\ref{fig:gemv_xeon}) saw similar gains. The reason for performance
improvement over simple unrolling is attributed to the fact that software
pipelining breaks dependency chains and provides the processor's ready
instruction window with a variety of independent instructions to choose from,
thus reducing the risk of a dependency stall.

%% NOTE: need to explain the delay thing better
Interestingly, software pipelining did not provide as significant of a gain on
the Intel Atom (Fig.~\ref{fig:gemv_atom}) as it did on the out-of-order
systems. This could be attributed to the fact that we were using this
processor in 32bit mode and thus we were limited to half the number of
registers that are available in 64bit mode. Furthermore, our current software
pipelining implementation does not support long delays, which might have been
necessary on this system.

@@

Additionally, for the sake of the in-order systems such as the Atom, or those
with lengthy instruction, we plan on implementing some mechanism for delaying
instructions for arbitrary lengths.

>>>>

\begin{abstract}
   Certain classes of algorithms are still hard to optimize by modern compilers
   due to the differences between computer architectures.
   Numerical algorithms, for example, perform better when taking advantage of features
   such as software pipelining and loop unrolling. Expert programmers are usually able to
   get the last drop of performance from the hardware, however, compilers cannot perform the same
   feat for loss of generality. The SPIRAL system bridges this gap by generating machine specific code
   from high level algorithmic descriptions. We used the SPIRAL system to implement software
   pipelining for numerical algorithms in order to bridge this gap. Our results show that
   by unrolling and applying software pipelining at the C code level, we can
   improve performance by up to 30\% over unrolling alone.
\end{abstract}

@@

In this project we explore the problem of scheduling code at the high-level and generate high-performance code that is comparable
to an expert.

@@

\begin{description}
   \item[Software Pipelining] We developed a transform that takes a numerical algorithm and adds software pipelining support
   using three stages. We are able to use this optimization along with other optimizations such as loop unrolling and loop tilling.
   Moreover, we were able to apply a technique that disallows C compilers to perform their own scheduling to the code. At the same
   time, we still use all the other optimizations provided by the C compiler. 
   \item[Evaluation] We evaluated our programs using several tunable parameters for software pipelining,
   loop unrolling and loop tilling. The results show very good speedups over code that only uses loop unrolling.
\end{description}

@@

Software pipelining is a coding technique that optimizes loops by overlapping instructions
from different loop iterations in order to exploit instruction level parallelism.
Due to this overlap we say that software pipelining is a type of Out-Of-Order execution.
While software pipelining is somewhat similar to hardware pipelining, the order of the executions
is not determined by the processor but by the compiler or the programmer.

@@

Most modern compilers do not preserve instruction ordering imposed at the
input language level, and instead rely on general purpose methods such
as \emph{list scheduling} to order instructions. In the common case, this frees
the developer from the task of determining an efficient schedule, but in our
case this feature serves as a roadblock. While the GCC~\cite{gccwebsite} compiler provides flags for
disabling instruction scheduling it comes at the cost of overall
performance because we only plan to schedule loop code. Moreover, our plan
involves using the Intel Compiler~\cite{iccwebsite}, thus we need a general solution for
scheduling at the C code level.

@@

Through the ability to apply software pipelining on any loop in SPIRAL and a
mechanism to enforce scheduling we can tailor a schedule of any
operation written in ICode to a given system. Currently, our optimizations for
tailoring software pipelining include loop tilling (strip-mining), reordering and unrolling the
outer loops before we perform the actual scheduling. These few optimization
provide a great degree of freedom with respect to tuning an implementation to
a system.

However, our current system is not without limitations. Because we are
attempting to schedule at a high level, our ICode implementation must map to the
underlying Instruction Set Architecture (ISA). Furthermore, our method
restricts the quality of optimizations that the compiler can provide, thus the
ICode must be aggressively optimized. Fortunately, SPIRAL was designed with
the necessary facilities in place for this level of optimization.

@@

The goal of our experiments is to demonstrate the effectiveness of software
pipelining at the C code level for automatically generated numerical
operations. Instruction scheduling is one of the few optimizations that programmer experts
have over SPIRAL. Thus, showing that software pipelining in addition to loop
unrolling outperforms unrolling alone will help in bridging the gap between
experts and APGS like SPIRAL.

%% NOTE: need to explain this WAY better in the future.
For our experiments we benchmarked multiple automatically generated variants
of Matrix-Vector (GEMV) on several systems with different
microarchitectures. The variants included a baseline in which no loops were
unrolled, two unrolled variants where the inner-most loop was fully unrolled
and the other dimension was unrolled by either 1 or 4. Additionally, there
were two variants that were software pipelined versions of the unrolled
code. The two different unrollings resulted in completely different schedules
in the software pipelined code. It is worth noting that we specifically chose
our input sizes to fit entirely in the L1 cache, so that performance would
mostly be impacted by the number of instruction stalls. Lastly, we used
4-way 32bit floating point vector instructions.

%% NOTE: in future, need actual parameters on these systems.
The systems we tested on were selected for their varied scheduling
facilities with respect to instruction window size where applicable. We chose three
Out-Of-Order architectures (Core i7 960, Core2 X9650, Xeon E5320) and a single
in-order system (Intel Atom).

@@

As anticipated, the use of software pipelining provided a substantial
performance gain over unrolling alone, and on the more modern system (Core i7,
Fig.~\ref{fig:gemv_corei7}) yielded an impressive 1.8x speedup over unrolling
alone. The other two Out-Of-Order systems (Fig.~\ref{fig:gemv_core2} and
Fig.~\ref{fig:gemv_xeon}) saw similar gains. The reason for performance
improvement over simple unrolling is attributed to the fact that software
pipelining breaks dependency chains and provides the processor's ready
instruction window with a variety of independent instructions to choose from,
thus reducing the risk of a dependency stall.

%% NOTE: need to explain the delay thing better
Interestingly, software pipelining did not provide as significant of a gain on
the Intel Atom (Fig.~\ref{fig:gemv_atom}) as it did on the Out-Of-Order
systems. This could be attributed to the fact that we were using this
processor in 32bit mode and thus we were limited to half the number of
registers that are available in 64bit mode. Furthermore, our current software
pipelining implementation does not support long delays, which might have been
necessary on this system.

@@

Additionally, for the sake of the In-Order systems such as the Atom, or those
with lengthy instruction, we plan on implementing some mechanism for delaying
instructions for arbitrary lengths.
